Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index

### 2016-06-20 swsc.tex
Draft 2016/06/20

   \author{M. Chandorkar
          \inst{1}
          \and
          E. Camporeale\inst{1}
          \and 
          S. Wing \inst{1,2}
          }

@@

   \abstract{
    We present two models that improve the state of the art of \emph{One Step Ahead}(OSA) prediction of the $Dst$ geomagnetic activity index. In the domain of OSA prediction, it is well known that $Dst(t)$ values correlate strongly with values one hour in the past (i.e. $Dst(t-1)$). While a number of forecasting models have been proposed and compared, none of them have been compared with the \emph{Persistence} model to quantify their performance improvements above this most trivial of predictors. In this study we train Gaussian Process regression models for One Step Ahead (OSE) prediction of the Disturbance Storm Time ($Dst$) geomagnetic index. We propose two variants of the Gaussian Process model.
    \begin{enumerate*}
      \item Gaussian Process Auto regressive (GP-AR) 
      \item Gaussian Process Auto regressive with eXogenous inputs (GP-ARX). 
    \end{enumerate*}
    We compare the performance of these models with the current state of the art in one step ahead $Dst$ prediction on a set of 
    63 benchmark storms from 1998-2006. The proposed Gaussian Process models not only outperform the rest but also the \emph{Persistence} model.
    }

@@

Geomagnetic indices come in various forms, they may take continuous or discrete values, indices may be calculated at varying time resolutions. Their values are often calculated by averaging or combining a number of readings taken by instruments around the Earth. Each geomagnetic index is a proxy for a particular kind of phenomena. Some popular indices are the $K_p$, $Dst$ and the $AE$ index.

\begin{enumerate}
    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices \cite{Bartels}. The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.

    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \cite{AEIndex}. It is a continuous index which is calculated every hour.

    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents and the strength of geomagnetic storms \cite{DesslerAndParker}. 
\end{enumerate}

%Talk about Burton and friends
For the present study, we focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms. More specifically, we focus on the \emph{one step ahead} (OSA), in this case one hour ahead prediction of $Dst$ because it is the first piece towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature. One the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where Q(t) relates to the particle injection from the plasma sheet into the inner magnetosphere.

@@

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated using the standard \emph{ordinary least squares} (OLS) method to minimize an objective function. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

%Talk about neural networks
Yet another family of forecasting methods is based on \emph{Artificial Neural Networks} (ANN) that have been a popular choice for building predictive models. Researchers have employed both the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures. \citet{Lund} proposed an \emph{Elman} recurrent network architecture called Lund $Dst$, which used the solar wind velocity, \emph{interplanetary magnetic field} (IMF) and historical $Dst$ data as inputs. \citet{JGRA:JGRA17461} used the recurrent neural networks to predict $Kp$. \citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} [see \citet{boyle1997empirical}]. The same architecture is adapted for prediction of $Dst$ in \citet{SWE:SWE286}, which is popularly known as the Rice $Dst$ model. \citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $Dst$ using only the IMF data.

%Talk about comparison papers, number of storms covered and Persistence
In light of the extensive list of modeling techniques employed for prediction of the $Dst$ index, model comparison and evaluation becomes a crucial direction in advancing the research domain. \citet{Rastatter2013} compare several physics based (convection, kinetic and magneto hydrodynamic) and empirical prediction models such as NARMAX, Rice $Dst$ on 4 storm events which occurred between 2001 and 2006. \citet{Amata2008} compare the EDDA and the Lund $Dst$ models over the 2003-2005 period. The most extensive model comparison in terms of storm events was probably conducted in \citet{Ji2012} which compared six $Dst$ models (see table \ref{table:DstModels}) on a list of 63 geomagnetic storm events of varying intensities which occurred between 1998 and 2006. In the comparison done in \citet{Ji2012}, the model proposed in \citet{JGRA:JGRA16300} gives the best performance on the test set considered.

%Vision
In $Dst$ prediction, a seemingly trivial yet highly informative prediction method is represented by the so called \emph{Persistence} model, which uses the previous value of $Dst$ as the prediction for the next time step ($\hat{D}st(t) = Dst(t-1)$). Due to high correlation between $Dst$ values one hour apart ($Dst(t)$, $Dst(t-1)$), despite its apparent lack of sophistication the \emph{Persistence} model gives excellent predictive performance on error metrics. In essence it is a trivial hypothesis and hence it should always be used as a base line to compare the performance of any proposed $Dst$ algorithm. Moreover, given its zero computational cost, a model should at least outperform the \emph{Persistence} model in order to be a viable candidate for advancing the science of space weather prediction.

%Assertion
In this paper, we propose \emph{Gaussian Process} models for OSA prediction of $Dst$. We use the results of \citet{Ji2012} as a starting point and compare our proposed models with the methods outlined in the same paper while using the performance of the \emph{Persistence} model as a base line.

% Talk about Gaussian Process
We use the hourly resolution data as extracted from NASA/GSFC's OMNI data set through OMNIWeb (see \cite{OmniPaper}). The rest of the paper is organized as follows. Section \ref{sec:method} gives an introduction to the \emph{Gaussian Process} methodology for regression and some technical details on their application. In section \ref{sec:osa} two different \emph{Gaussian Process} auto-regressive models are proposed and subsequently in sections \ref{sec:exp} and \ref{sec:res}, bench marked against the \emph{Persistence} $Dst$ model as well as the other models outlined in \citet{Ji2012}.

@@

On the basis of a set of labeled data points ${(\mathbf{x}_i, y_i); i = 1 \cdots N}$ which can be conveniently expressed by $\mathbf{X}$ a $N \times d$ data matrix and $\mathbf{y}$ a $N \times 1$ response vector as shown in equations \ref{eq:feat} and \ref{eq:labels}.

@@

Out task is to infer the the values of the unknown function $f(.)$ based on the inputs $\mathbf{X}$ and the noisy observations $\mathbf{y}$. We now assume that the joint distribution of $f(\mathbf{x}_i), i = 1 \cdots N$ is multivariate Gaussian as shown in equations \ref{eq:fvalues}, \ref{eq:normal} and \ref{eq:sto}.

@@

Here $\mathbf{f}$ is an $N\times 1$ vector consisting of the values $f(\mathbf{x}_i), i = q \cdots N$. In equation \ref{eq:normal} $\mathbf{f}|\mathbf{x}_1, \cdots, \mathbf{x}_N$ denotes the conditional distribution of $\mathbf{f}$ with respect to the input data (i.e. $\mathbf{X}$) and $\mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)$ represents a multivariate Gaussian distribution with mean vector $\mathbf{\mu}$ and covariance matrix $\mathbf{\Lambda}$. The probability density function of this distribution $p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N)$ is therefore given by equation \ref{eq:sto}.

From equation \ref{eq:sto}, one can observe that in order to uniquely define the distribution of the process, it is required to specify $\mathbf{\mu}$ and $\mathbf{\Lambda}$. For this probability density to be valid, there is a further requirements imposed on $\mathbf{\Lambda}$:

\begin{enumerate}
      \item Symmetry: $\mathbf{\Lambda}_{ij} = \mathbf{\Lambda}_{ij} \ \forall i,j \in {1, \cdots, N} $ 
      \item Positive Semi-definiteness: $\mathbf{z}^T \mathbf{\Lambda} \mathbf{z} \geq 0 \ \forall \mathbf{z} \in \mathbb{R}^N$  
\end{enumerate}

@@

Here $\mathbb{E}$ denotes the expectation (average). The elements of $\mathbf{\mu}$ and $\mathbf{\Lambda}$ are expressed as functions $m(\mathbf{x}_i)$ and $K(\mathbf{x}_i, \mathbf{x}_j)$ of the inputs $\mathbf{x}_i,\ \mathbf{x}_j$. Specifying the functions $m(\mathbf{x})$ and $K(\mathbf{x}, \mathbf{x}')$ completely specifies each element of $\mathbf{\mu}$ and $\mathbf{\Lambda}$ and subsequently the finite dimensional distribution of $\mathbf{f} | \mathbf{x}_1, \cdots, \mathbf{x}_N $. In most practical applications of \emph{Gaussian Processes} the mean function is often defined as $m(\mathbf{x}) = 0$, which is not unreasonable if the data is standardized to have zero mean. \emph{Gaussian Processes} are represented in machine learning literature using the following notation.

@@

Our aim is to infer the function $f(\mathbf{x})$ from the noisy training data and generate predictions $f(\mathbf{x}^{*}_i)$ for a set of test points $ {\mathbf{x}^{*}_i : \forall i \in 1, \cdots, M} $. We define $\mathbf{X}^*$ as the test data matrix whose rows are formed by $\mathbf{x}^{*}_i$ as shown in equation \ref{eq:testfeat}. Using the multivariate Gaussian distribution in equation \ref{eq:sto} we can construct the distribution of $f(\mathbf{x})$ over the training and test points. The vector of training and test outputs $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right)$ is of dimension $(N+M) \times 1$ and is constructed by appending the test set predictions $\mathbf{f}_*$ to the observed noisy measurements $\mathbf{y}$.

\begin{eqnarray}
    \mathbf{X}_* & = & \left( \begin{array}{c} (\mathbf{x}^{*}_1)^T \\ (\mathbf{x}^{*}_2)^T \\ \vdots \\ (\mathbf{x}^{*}_M)^T \end{array} \right)_{M \times d} \label{eq:testfeat} \\
    \mathbf{f}_* & = & \left( \begin{array}{c} f(\mathbf{x^{*}_1}) \\ f(\mathbf{x^{*}_2}) \\ \vdots \\ f(\mathbf{x^{*}_M}) \end{array} \right)_{M \times 1} \\
     \vspace{4\baselineskip}
    \left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_* & \sim & 
    \mathcal{N}\left(\mathbf{0}, \left[ \begin{array}{cc} \mathbf{K} + \sigma^{2} \mathbf{I} & \mathbf{K}_{*} \\ \mathbf{K}_{*}^T & \mathbf{K}_{**} \end{array} \right ] \right) \label{eq:dist}
\end{eqnarray}

Since we have noisy measurements of $f$ over the training data, we add the noise variance $\sigma^2$ to the variance of $f$ as shown in \ref{eq:dist}. The block matrix components of the $(N+M) \times (N+M)$ covariance matrix have the following structure.

\begin{enumerate}
      \item $\mathbf{I}$: The $n \times n$ identity matrix.
      \item $\mathbf{K} = [K(\mathbf{x}_i, \mathbf{x}_j)], \ i,j \in 1,\cdots,n$ : Kernel matrix constructed from all couples of the training data.
      \item $\mathbf{K}_{*} = [K(\mathbf{x}_i, \mathbf{x}^{*}_j)], \ i \in 1,\cdots,n ; j \in 1,\cdots,m$ : Cross kernel matrix constructed from all couples between training and test data points.
      \item $\mathbf{K}_{**} = [K(\mathbf{x}^{*}_i, \mathbf{x}^{*}_j)], \ i,j \in 1,\cdots,m$: Kernel matrix constructed from all couples of the test data.
\end{enumerate}

With the multivariate normal distribution defined in equation \ref{eq:dist}, probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. Since the original distribution of $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_*$ is a multivariate Gaussian, conditioning on a subset of elements $\mathbf{y}$ yields another Gaussian distribution whose mean and covariance can be calculated exactly in equation \ref{eq:posterior} (see \citet{Rasmussen:2005:GPM:1162254}).

\begin{eqnarray}
    \mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*} & \sim & \mathcal{N}(\mathbf{\bar{f}_*}, \Sigma_*)  \label{eq:posterior} \\
    \mathbf{\bar{f}_*} & = & \mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y} \label{eq:posteriormean} \\
    \Sigma_* & = & \mathbf{K}_{**} - \mathbf{K}^T_{*} \left(\mathbf{K} + \sigma^{2} \mathbf{I}\right)^{-1} \mathbf{K}_{*} \label{eq:posteriorcov}
\end{eqnarray}

The practical implementation of \emph{Gaussian Process} models requires the inversion of the training data kernel matrix $[\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1}$ to calculate the parameters of the predictive distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. The computational complexity of this inference is dominated by the linear problem $\mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y}$, this can be done via Cholesky decomposition which has time complexity $O(N^3)$.

The distribution of $\mathbf{f_*}| \mathbf{X},\mathbf{y},\mathbf{X_*}$ is known in Bayesian analysis as the \emph{Posterior Predictive Distribution}. This illustrates a key difference between \emph{Gaussian Processes} and other regression models such as \emph{Neural Networks}, \emph{Linear Models} and \emph{Support Vector Machines}, that a \emph{Gaussian Process} model does not generate point predictions for new data but outputs a predictive distribution for the quantity sought, this allows us to construct error bars on the predictions. This property of Bayesian models such as \emph{Gaussian Processes} makes them suitable for Space Weather forecasting applications.

The central design issue in applying \emph{Gaussian Process} models is choice of the function $K(\mathbf{x}, \mathbf{x}')$. The same constraints that apply to $\mathbf{\Lambda}$ also apply to the function $K(.,.)$. In machine learning, these symmetric positive definite functions of two variables are known as \emph{kernels}. Kernel based methods are applied extensively in data analysis i.e. regression, clustering, classification, density estimation (see \citet{Scholkopf:2001:LKS:559923}, \citet{hofmann2008}).

@@

For the success of a Gaussian Process model an appropriate choice of kernel function is paramount. The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data. The interested reader is suggested to refer to \cite{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment on kernel functions and the rich theory behind them. Some common kernel functions used in machine learning include the radial basis function (RBF) kernel $K(\mathbf{x}, \mathbf{y}) = \frac{1}{2} exp(-||\mathbf{x} - \mathbf{y}||^2/l^2)$ and the polynomial kernel $K(\mathbf{x}, \mathbf{y}) = (\mathbf{x}^\intercal \mathbf{y} + b)^d$.

The quantities $l$ in the RBF, $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}. Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The method of assigning values to the \emph{hyper-parameter} is crucial in the model building process.

In this study, we construct Gaussian Process regression models with linear kernels as shown in equation \ref{eq:usedKernel}, for \emph{one step ahead} prediction of the $Dst$ index. Our choice of kernel leaves us with only one adjustable hyper-parameter, the model noise $\sigma$.

@@

We initialize a grid of values for $\sigma$ the model noise and use the error on a predefined validation data set to choose the best performing value of $\sigma$. While constructing the grid of possible values for $\sigma$ it must be ensured that $\sigma > 0$ so that the kernel matrix constructed on the training data is non-singular.

@@

\subsection{Future Research Directions}

>>>>

   \author{M. Chandorkar
          \inst{1}
          \and
          E. Camporeale\inst{1}
          \and 
          S. Wing \inst{2}
          }

@@

   \abstract{
    We present two models, based on Gaussian Processes, for the \emph{One Step Ahead}(OSA) prediction of the $Dst$ geomagnetic activity index. 
    The models are Auto regressive, with or without eXogenous inputs (GP-ARX and GP-AR, respectively).
    We compare the performance of these models with the current state of the art in one step ahead $Dst$ prediction on a set of 
    63 benchmark storms from 1998-2006, previously analyzed in an earlier study.
    We show that, despite its lack of sophistication, the so-called Persistence model represents an important test to evaluate the performance of a one hour ahead $Dst$ model.
    Contrary to the state-of-the-art models compared in the literature, our models consistently outperform the Persistence model and represent a substantial improvement in the field, when evaluated on standard metrics.
    Finally, an important feature of the new models is that they naturally provide confidence intervals for their forecast.
    }

@@

Geomagnetic indices come in various forms, they may take continuous or discrete values and may be defined with varying time resolutions. Their values are often calculated by averaging or combining a number of readings taken by instruments around the Earth. Each geomagnetic index is a proxy for a particular kind of phenomenon. Some popular indices are the $K_p$, $Dst$ and the $AE$ index.

\begin{enumerate}
    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices \citep{Bartels}. The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.

    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citep{AEIndex}. It is a continuous index which is calculated every hour.

    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents and the strength of geomagnetic storms \citep{DesslerAndParker}. 
\end{enumerate}

%Talk about Burton and friends
For the present study, we focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms. More specifically, we focus on the \emph{one step ahead} (OSA), in this case one hour ahead prediction of $Dst$ because it is the simplest model towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature. One of the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where Q(t) relates to the particle injection from the plasma sheet into the inner magnetosphere.

@@

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective objective function. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

%Talk about neural networks
Yet another family of forecasting methods is based on \emph{Artificial Neural Networks} (ANN) that have been a popular choice for building predictive models. Researchers have employed both the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures. \citet{Lund} proposed an \emph{Elman} recurrent network architecture called Lund $Dst$, which used the solar wind velocity, \emph{interplanetary magnetic field} (IMF) and historical $Dst$ data as inputs. \citet{JGRA:JGRA17461} used recurrent neural networks to predict $Kp$. \citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} \citep{boyle1997empirical}. The same architecture is adapted for prediction of $Dst$ in \citet{SWE:SWE286}, popularly known as the Rice $Dst$ model. \citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $Dst$ using only the IMF data.

%Talk about comparison papers, number of storms covered and Persistence
In light of the extensive list of modeling techniques employed for prediction of the $Dst$ index, model comparison and evaluation becomes a crucial step for advancing the research domain. \citet{Rastatter2013} compared several physics based (convection, kinetic and magneto hydrodynamic) and empirical prediction models such as NARMAX, Rice $Dst$ on 4 storm events which occurred between 2001 and 2006. \citet{Amata2008} compared the EDDA and the Lund $Dst$ models over the 2003-2005 period. The most extensive model comparison in terms of storm events was probably conducted in \citet{Ji2012} which compared six $Dst$ models (see table \ref{table:DstModels}) on a list of 63 geomagnetic storm events of varying intensities which occurred between 1998 and 2006. In the comparison done in \citet{Ji2012}, the model proposed in \citet{JGRA:JGRA16300} gives the best performance on the test set considered.

%Vision
In $Dst$ prediction, a seemingly trivial yet highly informative prediction method is represented by the so called \emph{Persistence} model, which uses the previous value of $Dst$ as the prediction for the next time step ($\hat{D}st(t) = Dst(t-1)$). Due to high correlation between $Dst$ values one hour apart ($Dst(t)$, $Dst(t-1)$), the \emph{Persistence} model gives excellent predictive performance on error metrics, despite its apparent lack of sophistication. In essence it is a trivial hypothesis and hence it should always be used as a base line to compare the performance of any proposed $Dst$ algorithm. Moreover, given its zero computational cost, a model should at least outperform the \emph{Persistence} model in order to be a viable candidate for advancing the science of space weather prediction. An obvious critique to the Persistence model is that it is not really predictive, in the sense that it cannot forecast a storm, until it has already commenced. However, most of the literature has used global metrics, such as the Root Mean Square Error (RMSE, see section \ref{sec:exp}) to evaluate models. It is in this context that we argue that the Persistence model should be regarded as the first candidate to outperform.

%Assertion
In this paper, we propose \emph{Gaussian Process} models for OSA prediction of $Dst$. We use the results of \citet{Ji2012} as a starting point and compare our proposed models with the methods evaluated in that paper, while using the performance of the \emph{Persistence} model as a base line.

% Talk about Gaussian Process
We use hourly resolution measurements extracted from NASA/GSFC's OMNI data set through OMNIWeb \citep{OmniPaper}. The rest of the paper is organized as follows. Section \ref{sec:method} gives an introduction to the \emph{Gaussian Process} methodology for regression and some technical details about their application. In section \ref{sec:osa} two different \emph{Gaussian Process} auto-regressive models are proposed and subsequently in sections \ref{sec:exp} and \ref{sec:res}, bench marked against the \emph{Persistence} as well as the other models outlined in \citet{Ji2012}.

@@

A set of labeled data points ${(\mathbf{x}_i, y_i); i = 1 \cdots N}$ can be conveniently expressed by a $N \times d$ data matrix $\mathbf{X}$ and a $N \times 1$ response vector $\mathbf{y}$, as shown in equations (\ref{eq:feat}) and (\ref{eq:labels}).

@@

Our task is to infer the values of the unknown function $f(.)$ based on the inputs $\mathbf{X}$ and the noisy observations $\mathbf{y}$. We now assume that the joint distribution of $f(\mathbf{x}_i), i = 1 \cdots N$ is a multivariate Gaussian as shown in equations (\ref{eq:fvalues}), (\ref{eq:normal}) and (\ref{eq:sto}).

@@

Here $\mathbf{f}$ is a $N\times 1$ vector consisting of the values $f(\mathbf{x}_i), i = 1 \cdots N$. In equation (\ref{eq:normal}), $\mathbf{f}|\mathbf{x}_1, \cdots, \mathbf{x}_N$ denotes the conditional distribution of $\mathbf{f}$ with respect to the input data (i.e., $\mathbf{X}$) and $\mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)$ represents a multivariate Gaussian distribution with mean vector $\mathbf{\mu}$ and covariance matrix $\mathbf{\Lambda}$. The probability density function of this distribution $p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N)$ is therefore given by equation (\ref{eq:sto}).

From equation (\ref{eq:sto}), one can observe that in order to uniquely define the distribution of the process, it is required to specify $\mathbf{\mu}$ and $\mathbf{\Lambda}$. For this probability density to be valid, there are further requirements imposed on $\mathbf{\Lambda}$:

\begin{enumerate}
      \item Symmetry: $\mathbf{\Lambda}_{ij} = \mathbf{\Lambda}_{ji} \ \forall i,j \in {1, \cdots, N} $ 
      \item Positive Semi-definiteness: $\mathbf{z}^T \mathbf{\Lambda} \mathbf{z} \geq 0 \ \forall \mathbf{z} \in \mathbb{R}^N$  
\end{enumerate}

@@

Here $\mathbb{E}$ denotes the expectation (average). The elements of $\mathbf{\mu}$ and $\mathbf{\Lambda}$ are expressed as functions $m(\mathbf{x}_i)$ and $K(\mathbf{x}_i, \mathbf{x}_j)$ of the inputs $\mathbf{x}_i,\ \mathbf{x}_j$. Specifying the functions $m(\mathbf{x})$ and $K(\mathbf{x}, \mathbf{x}')$ completely specifies each element of $\mathbf{\mu}$ and $\mathbf{\Lambda}$ and subsequently the finite dimensional distribution of $\mathbf{f} | \mathbf{x}_1, \cdots, \mathbf{x}_N $. In most practical applications of \emph{Gaussian Processes} the mean function is often defined as $m(\mathbf{x}) = 0$, which is not unreasonable if the data is standardized to have zero mean. \emph{Gaussian Processes} are represented in machine learning literature using the following notation:

@@

Our aim is to infer the function $f(\mathbf{x})$ from the noisy training data and generate predictions $f(\mathbf{x}^{*}_i)$ for a set of test points $ {\mathbf{x}^{*}_i : \forall i \in 1, \cdots, M} $. We define $\mathbf{X}^*$ as the test data matrix whose rows are formed by $\mathbf{x}^{*}_i$ as shown in equation (\ref{eq:testfeat}). 
\begin{equation}
    \mathbf{X}_* = \left( \begin{array}{c} (\mathbf{x}^{*}_1)^T \\ (\mathbf{x}^{*}_2)^T \\ \vdots \\ (\mathbf{x}^{*}_M)^T \end{array} \right)_{M \times d} \label{eq:testfeat} 
\end{equation}
Using the multivariate Gaussian distribution in equation (\ref{eq:sto}) we can construct the joint distribution of $f(\mathbf{x})$ over the training and test points. The vector of training and test outputs $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right)$ is of dimension $(N+M) \times 1$ and is constructed by appending the test set predictions $\mathbf{f}_*$ to the observed noisy measurements $\mathbf{y}$.

\begin{eqnarray}
    \mathbf{f}_* & = & \left( \begin{array}{c} f(\mathbf{x^{*}_1}) \\ f(\mathbf{x^{*}_2}) \\ \vdots \\ f(\mathbf{x^{*}_M}) \end{array} \right)_{M \times 1} \\
     \vspace{4\baselineskip}
    \left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_* & \sim & 
    \mathcal{N}\left(\mathbf{0}, \left[ \begin{array}{cc} \mathbf{K} + \sigma^{2} \mathbf{I} & \mathbf{K}_{*} \\ \mathbf{K}_{*}^T & \mathbf{K}_{**} \end{array} \right ] \right) \label{eq:dist}
\end{eqnarray}

Since we have noisy measurements of $f$ over the training data, we add the noise variance $\sigma^2$ to the variance of $f$ as shown in (\ref{eq:dist}). The block matrix components of the $(N+M) \times (N+M)$ covariance matrix have the following structure.

\begin{enumerate}
      \item $\mathbf{I}$: The $n \times n$ identity matrix.
      \item $\mathbf{K} = [K(\mathbf{x}_i, \mathbf{x}_j)], \ i,j \in 1,\cdots,n$ : Kernel matrix constructed from all couples obtained from the training data.
      \item $\mathbf{K}_{*} = [K(\mathbf{x}_i, \mathbf{x}^{*}_j)], \ i \in 1,\cdots,n ; j \in 1,\cdots,m$ : Cross kernel matrix constructed from all couples between training and test data points.
      \item $\mathbf{K}_{**} = [K(\mathbf{x}^{*}_i, \mathbf{x}^{*}_j)], \ i,j \in 1,\cdots,m$: Kernel matrix constructed from all couples obtained from the test data.
\end{enumerate}

With the multivariate normal distribution defined in equation (\ref{eq:dist}), probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. Since the original distribution of $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_*$ is a multivariate Gaussian, conditioning on a subset of elements: $\mathbf{y}$ yields another Gaussian distribution whose mean and covariance can be calculated exactly, as in equation (\ref{eq:posterior}) (see \citet{Rasmussen:2005:GPM:1162254}).
\begin{equation}
    \mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*} \sim \mathcal{N}(\mathbf{\bar{f}_*}, \Sigma_*)  \label{eq:posterior},
\end{equation}
where
\begin{eqnarray}
    \mathbf{\bar{f}_*} & = & \mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y} \label{eq:posteriormean} \\
    \Sigma_* & = & \mathbf{K}_{**} - \mathbf{K}^T_{*} \left(\mathbf{K} + \sigma^{2} \mathbf{I}\right)^{-1} \mathbf{K}_{*} \label{eq:posteriorcov}
\end{eqnarray}

The practical implementation of \emph{Gaussian Process} models requires the inversion of the training data kernel matrix $[\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1}$ to calculate the parameters of the predictive distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. The computational complexity of this inference is dominated by the linear problem $\mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y}$, which can be solved via Cholesky decomposition, with a time complexity of $O(N^3)$, where $N$ is the number of data points.

The distribution of $\mathbf{f_*}| \mathbf{X},\mathbf{y},\mathbf{X_*}$ is known in Bayesian analysis as the \emph{Posterior Predictive Distribution}. This illustrates a key difference between \emph{Gaussian Processes} and other regression models such as \emph{Neural Networks}, \emph{Linear Models} and \emph{Support Vector Machines}: a \emph{Gaussian Process} model does not generate point predictions for new data but outputs a predictive distribution for the quantity sought, thus allowing to construct error bars on the predictions. This property of Bayesian models such as \emph{Gaussian Processes} makes them very appealing for Space Weather forecasting applications.

The central design issue in applying \emph{Gaussian Process} models is the choice of the function $K(\mathbf{x}, \mathbf{x}')$. The same constraints that apply to $\mathbf{\Lambda}$ also apply to the function $K$. In machine learning, these symmetric positive definite functions of two variables are known as \emph{kernels}. Kernel based methods are applied extensively in data analysis i.e. regression, clustering, classification, density estimation (see \citet{Scholkopf:2001:LKS:559923}, \citet{hofmann2008}).

@@

For the success of a \emph{Gaussian Process} model an appropriate choice of kernel function is paramount. The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data. The interested reader is suggested to refer to \cite{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them. Some common kernel functions used in machine learning include the radial basis function (RBF) kernel $K(\mathbf{x}, \mathbf{y}) = \frac{1}{2} exp(-||\mathbf{x} - \mathbf{y}||^2/l^2)$ and the polynomial kernel $K(\mathbf{x}, \mathbf{y}) = (\mathbf{x}^\intercal \mathbf{y} + b)^d$.

The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}. Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The method of assigning values to the \emph{hyper-parameters} is crucial in the model building process.

In this study, we construct Gaussian Process regression models with linear kernels as shown in equation (\ref{eq:usedKernel}), for \emph{one step ahead} prediction of the $Dst$ index. Our choice of kernel leaves us with only one adjustable hyper-parameter, the model noise $\sigma$.

@@

We initialize a grid of values for the model noise $\sigma$ and use the error on a predefined validation data set to choose the best performing value of $\sigma$. While constructing the grid of possible values for $\sigma$ it must be ensured that $\sigma > 0$ so that the kernel matrix constructed on the training data is non-singular.

@@

%\subsection{Future Research Directions}
